---
title: "R Practice"
author: "DSSB"
date: "October 20, 2016"
output: pdf_document
---

What are chunks?

"R code chunks can be used as a means render R output into documents or to simply display code for illustration.""

For the sake of our workshop, we will be mainly using them as a way to store our code in a clear manner.

Some helpful shortcuts for RMD Files:

Inserting Chunks:
ctrl + alt + i

Running Chunks: 
ctrl + alt + c

Running lines:
ctrl + enter while highlighting code that needs to be run

```{r}
character_variable = "A string value"
```

What did our code do?
We assigned a string to a variable named "character_variable".

What types of data types are there?
1. Vectors
2. Matrices
3. Arrays
4. Data Frames
5. Lists

Let's start with Vectors:

What types of vectors are there?
There are 3: numerical, character, logical

```{r numerical_vectors}
num_vec_1 = c(10, 29102, 6.954, -2.2)
num_vec_2 = c(1:4)
num_vec_3 = as.vector(0:10)
```

```{r character_vectors}
char_vec_1 = c("a", "test,", "to", "check")
char_vec_2 = c("is this.", "Character")
```

```{r logical_vectors}
logic_vec_1 = c(TRUE, FALSE, TRUE) 
logic_vec_2 = c(T, F, T, F)
```

```{r vector_questions}
#Q1: How can I create a numerical vector containing the integers 50 through 100 inclusively?

#Q2: How can I create a character vector that says TRUE?

```


```{r vector_operations}
num_vec_1 + num_vec_2 #num + num
sum(num_vec_2, num_vec_3) #num + num

num_vec_2 + logic_vec_2 #num + logic

num_vec_2 + char_vec_1 #num + char
logic_vec_2 + char_vec_1 #num + char 

num_vec_2 * 2 

max(num_vec_2)
min(num_vec_2)
head(char_vec_2, n=2)
tail(logic_vec_2, n= 2)

```

```{r vector_op_qs}
#Q1: What should be the result of logic_vec_2 + num_vec_1

#Q2: What should be the result of 3 * logic_vec_1 

#Q3: What is the max value of num_vec_1?
```


```{r vector_operations_2}
length(num_vec_1)
str(num_vec_1)

sort(num_vec_1) #returns sorted vector of numerics
sort(num_vec_1, decreasing=TRUE) #returns sorted vector of numerics in decreasing pattern

order(num_vec_1, decreasing=FALSE) #orders the vector by indices
num_vec_1[order(num_vec_1, decreasing=FALSE)] #spits out ordered vector by indices
```

```{r vector_slicing}
num_vec_1[3]
num_vec_1[5] #Why is the output this way?

num_vec_1[2:4]
num_vec_1[-2:-1]

#Q: How can I obtain the 3rd to 6th elements of num_vec_3?

```

Lists are different from vectors: lists will not try to convert all types of entries into a certain type.

```{r lists}
lst_1 = list("word", 1, c(4, 5), T)
lst_1
length(lst_1)

lst_2 = list(lst_1, c("a", "b", 5))
```

How can we extract certain entries from a list?
```{r lists_operations}
lst_1[2] #by entry
lst_1[2:4] #by subset
```

Matrices & Arrays:

All columns in a matrix must have the same mode(numeric, character, etc.) and the same length
```{r}
num_matrix = matrix(
  c(1:8),
  nrow=2,
  ncol=4,
  byrow = TRUE #Q: what happens if byrow line is gone?
)

char_matrix = matrix(
  c("A", "Small", "Exercise", "To", "Practice", "R"),
  nrow=3,
  ncol=2,
  byrow = FALSE
)

#Extracting values
char_matrix[2, 2]
char_matrix[1, ]
char_matrix[, 2]
```

Arrays are similar to matrices, but can have more than two dimensions
```{r}
array_vec_1 = c(5, 6, 8)
array_vec_2 = c(10:15)
num_array = array(c(array_vec_1, array_vec_2), dim=c(2, 3, 3))
num_array #Should be an array of three 2x3 matrices each
```

Tables:
```{r}
practice_table = table(char_vec_1, num_vec_2, logic_vec_2)
practice_table
nrow(practice_table)
str(practice_table)
```


Data Frames: 

Very, very useful for containing data, as they can hold data from any type of mode. 
```{r}
df_col_1 = c(1:5)
df_col_2 = c("A", "B", "C", "D", "E")
df_col_3 = c(T, F, T, F, T)

practice_df = data.frame(df_col_1, df_col_2) #create data frame with 2 columns
practice_df$df_col_3 = df_col_3 #add a column

practice_df

class(practice_df[2, 1])
class(practice_df[3, 3])
```

Here are some basic things we can do with data frames.

```{r df_subset_rows}
practice_df[2:4, 1:2] #subsetting indices
practice_df$df_col_3 # $ notation
practice_df[practice_df$df_col_2 == "A", ] #subsetting conditionals
practice_df[practice_df$df_col_3 == T | practice_df$df_col_1 == 4, ]

practice_df[practice_df$df_col_1 > practice_df$df_col_3, ]
```

What else can we do?

For loops: typically not used often, as vectorization in R allows for quick access and replication. Terminates when the list is fully iterated through.
```{r for_loops}
for (month in c("Jan", "Feb", "March", "April", "May", "June",
                "July", "Aug", "Sept", "Oct", "Nov", "Dec")){
  print(paste("The month is", month)) #similar to cat!
}
```

While loops: Similar to for loops, except the loop terminates when the condition is no longer met.
```{r while_loops}
while_count = 0

while (while_count < 10) {
  print("Did it work?")
  while_count = while_count + 1 #We have to increase the count to continue
}

#This could've been simplified as:
replicate("Did it work", n=10)
```

Functions: Handy to address our own custom issues.
```{r functions}
quadratic_formula = function(a, b, c){ 
  x_1 = ((-b) + sqrt(b^2 - 4*a*c))/(2*a)
  x_2 = ((-b) - sqrt(b^2 - 4*a*c))/(2*a)
  return(c(x_1, x_2))
}
quadratic_formula(1, 1, -2)
```

```{r functions_practice}
#Q: Write down the formula for calculating the midpoint between two points. Assume each point has an x-value and a y-value.

midpoint_formula = function(a, b){

}
midpoint_formula(c(2,4), c(4, 8))
```


Viewing Polling Data in Comparison to Tweets.

```{r poll_df}
poll_data = read.csv("hai.csv", stringsAsFactors = FALSE)
#Let's first make a copy of our data frame.
head(poll_data)
polls = poll_data

#Some basic exploration
colnames(polls)
row.names(polls)
str(polls)

#The first column is unnecessary so let's drop it.
polls = polls[-1]

#Let's add another column containing the numerical spread.
polls$Difference = abs(polls$Clinton - polls$Trump)

```

```{r df_basic_analysis}
#Let's check out how many times surveys have had Clinton in the lead versus Trump.
nrow(polls[polls$Clinton > polls$Trump, ])

#Let's do the same but with Trump in the lead instead.
nrow(polls[polls$Trump > polls$Clinton, ])

#Let's get the number of ties.
nrow(polls[polls$Trump == polls$Clinton, ])

#Let's see the highest margin a survey has placed Trump in the lead, and which survey conducted the survey.
trump_highest = max(polls$Trump - polls$Clinton)
polls[polls$Trump == polls$Clinton + trump_highest, ]

#Let's do the same but with Clinton. 
clinton_highest = max(polls$Clinton - polls$Trump)
polls[polls$Clinton == polls$Trump + clinton_highest, ]
```

## R Markdown

Let's grab the package we need to graph first.

```{r}
install.packages("ggplot2")
library("ggplot2")
```

Start by examining how the frequency of polls varies with date

```{r}
poll_plot = ggplot(data = polls)
poll_plot + geom_density(mapping = aes(x = Date)) 
```

Next, lets see how many people are usually sampled

```{r}
poll_plot + geom_boxplot(mapping = aes(x = 'density', y = Sample)) +
  scale_y_log10("Sample Size") 
```

Here we can look at general candidate approval over the past year and a half

```{r}
ggplot(data = polls, mapping = aes(x = Date)) + 
  geom_point(mapping = aes(y = Clinton), color = 'blue') +
  geom_smooth(mapping = aes(y = Trump), color = 'red') +
  scale_y_continuous('Candidate Approval') + theme(legend.position = "bottom")
```

